Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia.

### 2022-12-21 analysis/paper/paper.Rmd
- Commit after hard reset

---
title: "Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia."
author:
  - name: Joana Belmiro
    email: fl@oneplace.edu
    institute: [UofO]
    correspondence: true
  - name: João Cascalheira
    email: fl@another.edu
    institute: [UofA]
    correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

library(ggplot2)
theme_set(theme_bw())
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(readr)
library(ggspatial)
library(ggrepel)
library(dplyr)
library(kableExtra)
library(flextable)
library(data.table)
library(tidyr)

```{r databases, include=FALSE}

petrography <- read_csv(here::here('analysis/data/raw_data/PetrographyDB.csv')) # Reading the petrography data
chertfield <- read_csv(here::here('analysis/data/raw_data/chert_field.csv')) # Reading the field data

fielddata <- chertfield %>% # Cleaning the Field data
  mutate(Geography = case_when(long <= -8.2 ~ "Ocidental",
                               long > -8.2 ~ "Oriental")) %>% 
  filter(`Raw Material` == "Chert") %>% 
  rename(SampleID = NewCode)

```{r map setup, include=FALSE}

sf::sf_use_s2(FALSE)
world <- ne_countries(scale = "medium", returnclass = "sf")
#europe <- world[world$continent == "Europe",]
europe_cropped <- st_crop(world, xmin = -15, xmax = -7.5,
                                    ymin = 35, ymax = 37.6)

```{r map, echo=FALSE, include=FALSE}

ggplot(data = europe_cropped) +
  geom_sf() +
  geom_point(data = chertfield, aes(x = long, y = lat, shape = State, color = State), size = 4) +
  #geom_label_repel(data = chertsamples, aes(long, lat, label=NewCode), box.padding = 0.5, max.overlaps = 30) +
  annotation_scale(location = "br", width_hint = 0.5) +
  annotation_north_arrow(location = "br", which_north = "true", 
        pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
        style = north_arrow_fancy_orienteering)
  #coord_sf(xlim = c(long_min, long_max), ylim = c(lat_min,lat_max), expand = FALSE)

```{r databasesetup, include=FALSE}

CStable <- chertfield %>% 
  select(NewCode, Site, `Sample variability`, State, Deposit, Type, Frequence, `State of site`, `Thin-sections`)

```{r cstable, echo=FALSE, include=FALSE}

regulartable(CStable) %>% 
  theme_zebra()

@@

The South Iberian territory is a key area to understand cultural transitions from the Middle Paleolithic to early Prehistory . The Algarve region specifically (south Portugal) has contributed greatly to understanding pre-history at a European level, with several well-known archaeological sites and publications. As a territory located at the tip of the European continent and with a generally warm climate even during the coldest periods which characterize the Late Pleistocene, it has been regarded as one of the most significant glacial refugia in Europe (Hewitt, 2000; Gómez and Lunt, 2007; González-Sampériz et al. 2010).  Several authors have suggested that the Middle to Upper Paleolithic cultural sequences present in the Iberian Peninsula territory are related to adaptation and reorganization processes related to abrupt climatic events (Brädtmoller et al., 2012; Cascalheira et al., 2021). As such, the south Iberian territory witnessed several moments of crisis, cultural changes, and behavioral adaptations. An example of this is the late occupation of the territory by Neanderthal populations and their consequent disappearance.  The Iberian record for the Middle Paleolithic allows for nearly 250 kyr of Neanderthal occupations to be studied until the extinction interval (de La Torre, Martínez-Moreno, and Mora, 2013). Several studies suggest that Neanderthal presence in southern Iberia lasted until c. 37 ka cal BP or later, with a different trajectory than that found in other places of Europe (Cascalheira, Gonçalves, and Maio, 2021).

The several cultural shifts of the Upper Paleolithic (UP) are also represented in the region, in sites with long chronological occupations, from the first UP occupations such as the Aurignacian (not detected in southern Portugal) and the Gravettian, such as the archaeological site of Vale Boi . This site, containing one of the most complete UP sequences of the Iberian Peninsula, offers the opportunity to understand the movements of the first groups of Humans to reach the territory, their cultural changes throughout the Upper Paleolithic, as well as the impact of abrupt climatic change in culture, such as the Heinrich Event 2 (HE 2) and the Last Glacial Maximum (LGM). 
Finally, south Iberia also shows the existence of several Mesolithic, Neolithic, and Chalcolithic settlements and occupations. In south Portugal, for example, studies have shown the use of riverine and estuarine areas for Mesolithic settlements (Carvalho, 2008) and a concentration of sites by the westernmost coast of the Algarve (Bicho, 2009), as well as clear settlement and cultural shifts in the Neolithic occupations, followed by Chalcolithic occupations (Sánchez et al., 2012). This complete recent prehistory chronology makes southern Iberia, and in specific, southern Portugal, an ideal territory to study the replacement of the last hunter-gatherers of Europe by the first farmers, and the subsequent transformations these communities went through.

From at least ~40 000 cal BP to 3 000 cal BP (** , Valera, 2012), South Portugal has seen moments of crisis and cultural change of different kinds, which gives this territory a large potential for prehistorical archaeology studies and the study of human behavior in a wide chronological range. With such a wide chronology, differences in the exploitation of resources, management of territories, settlement patterns, subsistence techniques, climate, technology, and culture, it is important, however, to find a unifying subject that can be studied, analyzed, and compared, bringing together several studies and sites to create a more unified story of the evolution of human behavior. That subject may be the raw material that was used by Neanderthals, Human hunter-gatherers, and farmers - that raw material is chert.
Whenever geologically present (and at times, even when it is not geologically present), Chert has served as a raw material for the production of blanks to more specialized and complex tools. As a good quality rock, it was procured from close to far-away locations, transported, knapped, retouched, re-used, heated, and sometimes traded, traveling long distances in the hands of different groups which produced them. Countless archaeologists and studies have shown the potential of this raw material to understand cultural aspects such as foraging strategies and territory use, mobility, shifts in technological necessities, cultural preferences, changes in raw material visibility and accessibility, or even the establishment, expansion, and shrinkage of social networks ** . 
As a raw material, chert has been studied in South Portugal. The study of chert-bearing outcrops has been a topic of interest for both geologists and archaeologists in the Algarve region, albeit for different reasons. Geological literature has often focused on the processes that formed the seas and landmasses of the Algarve basins, as well as the paleobiology and paleobiogeography of the region. Archaeological literature has focused on chert sources as a way to predict the availability of mineral resources, and the existence of settlements, characterize hunter-gatherer culture throughout prehistory and understand group mobility and social exchange through time.
Geological studies in the Algarve region go back to the 19th century and throughout all of the 20th century, through the creation of geological cartography. These first works provided the basis for the geological knowledge of the Algarve basin, and it allowed further studies within the geosciences, as a large number of Doctoral theses.

Unlike what happens in other areas of Portugal (add references from Aubry, Abrunhosa, etc), archaeological studies focusing on raw materials, especially chert, have been more scarce. In general, lithic technology studies in the Algarve have approached raw materials through a macroscopic methodology, often related to the location of the settlement (i.e., [@bicho_o_2003]; Pereira et al 2016 ) or the description of the collection (i.e., [@cascalheira_tecnologia_2010], [@marreiros_lithic_2015]). The work of Veríssimo (2005) focusing on the occurrence of chert in western Algarve provided the initial basis for comparative studies, maintaining however solely a macroscopic component. Finally, the creation of LusoLit (Telmo et al. 2016) and the collection of samples from the Algarve provided a new leap in the study of chert in the region.
Given the existence of several chert-bearing outcrops in South Portugal, preliminary raw material studies, a region of archaeological importance, and the existence of a rich archaeological record, the Algarve stands as a region full of potential for studies of chert and the development of a lithotheque which may serve for regional studies, but also for comparative studies outside, where chert may have been transported and traded, offering the chance to complete the record of exotic cherts which may be coming from the Algarve.

This paper aims to locate and characterize the nature of chert sources in southern Portugal and their availability. Furthermore, it aims to test the potential of different methodological approaches to characterize the regional cherts and develop a methodological approach adapted to this study area.

@@

The Algarve region is located in southern Portugal, framed north and east by the Alentejo region and Spain respectively. To the west and south, it is bordered by the Atlantic ocean. Geologically, it is composed of two main geological units: the South Portuguese Zone (SPZ) and the Algarve basin. The SPZ is composed of four “domains” of different lithologies and ages, extending up to Alentejo, although all within the Paleozoic period. It is overlain unconformably by the Mesozoic sedimentary rocks of the basin [@fernandes_new_2012]. The Algarve basin is one of the broader Jurassic outcrops in Portugal (Malchus and Kuss, 1988). It corresponds to the Mesozoic-cenozoic sediments that outcrop south of Portugal, from Cape of St. Vicent to Vila Real de St. António. The Algarve basin is associated with the opening of the central Atlantic and with the eventual oceanic crust formation in the western part of the Tethys sea, between the Algarve and North Africa (Terrinha et al., 2013). After the sedimentation period between the Triassic and the Sinemurian (Lower Jurassic), the Mesozoic basin was divided into two sub-basins (western sub-basin and oriental sub-basin). The existence of two sub-basins and a structural higher area between them, allied with the expansion and retraction of the seas allowed a variation of sedimentation environments, ranging between the external platform and internal platform, and continental, hemipelagic, or deep marine environments (Terrinha et al., 2013). It was this variance in deposition environments that created the variety of Paleozoic sedimentary facies, with moments of more or less homogeneity throughout this period.
The Algarve basin is of essential importance for the existence of chert in the Algarve region since it is in the Paleozoic sediments where chert nodules can be found. Paleozoic sedimentation in the basin starts in the Triassic, although the identification of chert nodules or beds is limited to the Lower to Upper Jurassic. It is at the start of the Lower Jurassic, during the Lower Pliesbachian (also named Carixian) that the Algarve basin started to separate into two sub-basins until the Callovian (Middle Jurassic) when an uplift episode allowed uniform sedimentation of the basin (Rocha, 1976; Terrinha et al., 2013). During the Lower Pliensbachian, the sediments in the western sub-basin can be described as marine, of the external platform, while the sediments of the eastern sub-basin are of an internal platform. The transition between the Lower Jurassic to the Middle Jurassic shows a hiatus, as there are no known sediments from the Toarcian to the Aaelinian in the basin, the sedimentation restarting only after the latter age. During the Upper Jurassic, the lithofacies variation became more prominent, especially during the Upper Oxfordian-Lower Kimmeridgian, with a certain uniformity of the lithofacies in all the sub-basins during the Upper Kimmeridgian (Terrinha et al., 2013).

@@

In general, the presence of chert in the Algarve region and be attributed to limestone or dolomitic limestone layers. They appear with relative frequency, and often as nodules of small to medium dimensions, although thin beds of chert can also be identified in some areas.
As mentioned above, outcrops with chert nodules in the Algarve region can be found in the Paleozoic sediments from the Algarve Basin, more specifically in the Jurassic sediments. Chert nodules can be found within Lower, Middle and Upper Jurassic layers, although restricted to a limited number of layers within the stratigraphy and not present in every geological cut.

Lower Jurassic geological layers with chert nodules can be mostly found in the Carixian-age sediments (also known as Pliensbachian) from the Sagres region, often outcropping in those places where more layers are exposed, such as beach-generated cliffs and associated deposits. These include the chert nodules and beds from Cabo de S. Vicente to Praia do Belixe, but also the small outcrop inland, in Ferrel. The chert nodules occur within the limestone, dolomitic limestone, and marly limestone. Further east one layer with chert nodules was also identified but attributed to the Sinemurian-Toarcian, in a formation identified as “Limestones with Paleodasucladus and limestones with chert nodules”. Described only in the Geologic Map at a 1 200 000 scale, however, these chert nodules were characterized as micronodules [@oliveira_carta_1992].

Middle Jurassic geological layers with chert nodules are only found east of Sagres, in the Albufeira, Faro, and Tavira regions. These layers were attributed to the Malhão Formation, from the Aalenian-Bajocian age. In general, this formation was described as carbonated, marine, and thick, and chert can be found in two different layers: monogenic conglomerates with micritic limestone intercalations with chert beds and nodules; microcrystalline limestones with chert nodules (pel-biomicrite). Another chert outcrop was identified in the Geologic Map at a 1 200 000 scale, attributed to the Bajocian-Batonian age and named “Alagoa conglomerates”, with partially dolomitic limestones and chert nodules.

Finally, Upper Jurassic sediments have been identified also in the Albufeira, Faro, and Tavira area. The Jordana Formation was attributed to the Kimmeridgian/Kimmeridgian-Portlandian and is characterized by dark-gray limestones, with frequent secondary silifications with abundant fossil fragments. In the Tavira area, these layers were instead attributed to the Oxfordian-Kimmeridgian age.

@@

To identify the chert outcrops present in southern Portugal and understand the chert’s characteristics, a macroscopic and petrographic approach was applied to the study of geological samples. A combination of methods in a multilayered approach is an important tool to reconstruct the geological and geographical origin of raw materials. This is because different methods have their inherent limitations. By combining these two different methodologies the goal was to create several layers of analysis, which may then be further pursued or added new layers (Brandl, 2016). Several other similar methodologies and approaches have been applied in other regions (references). However, the chosen analysis techniques should be adapted to the specific geographic context, the research questions, problematics, and the characteristics of the types of cherts in question. Since only preliminary studies of raw materials were applied in the western portion of southern Portugal, and petrographic data has been shown to provide good results for the characterization of cherts in this region, the two methodologies were chosen for the study.
Geological samples were obtained through fieldwork, from August of 2021 (west Algarve) and April to June 2022 (east Algarve), and included visits to previously known outcrops and prospection of areas identified in geological maps. The first step of fieldwork included the review of previously known research, including geological maps, archaeological prospection works, and raw material studies (Pereira, Verissimo 2004, 2005). These were followed as a reference guide for recovering samples. Whenever necessary, samples from previous prospection works were consulted, from the LusoLit lithoteque from ICArEHB at the University of Algarve (Pereira) and the thin sections at the University of Évora. Whenever coordinates or specific locations for known outcrops were available, those were visited, and then the surrounding area was prospected to understand the extension of the outcrops and possible secondary deposition outcrops nearby. In all other situations, several locations with the potential to find outcrops were selected from polygons where formations with chert were known to the outcrop. 
Samples were collected whenever possible, focusing on both primary and secondary outcrops. When the outcrops showed macroscopic differences between the cherts, several samples were collected, to obtain as much variability as possible. All samples were georeferenced and registered with resources to an android app which allows accurate coordinates to be associated with photos and CSV data regarding the outcrop characteristics and conditions (i.e. abundance, visibility, access, geomorphology, chert morphology, and conditions). All data related to the app and dataset for the variables recorded can be found in the supplementary materials. Individual ID tags were associated with each sample, which was washed and labeled in the laboratory.
All geological samples  were analyzed macroscopically following a pre-established dataset. The variables were defined based on specialized literature and the dataset with the variables can be found in the Supplementary materials. The goal of this analysis was to describe the cherts and their enclosing rocks in a detailed and objective manner. For a comparative analysis of archaeological artifacts, other methods may be inconvenient or impossible to use, since they may be destructive and often difficult to apply to large assemblages. Macroscopic analyses have the advantage of being less costly and easy to apply. Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops, and formations is essential to apply the methodology with archaeological materials and establish comparisons (Bustillo et al 2009 ). A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with the resource of a Nikon SMZ25 stereomicroscope, focusing especially on aspects like inclusions and fossil content.
Thin sections were produced for geological samples of all formations, focusing on obtaining petrographic data which reflected the variability observed at a macroscopic level. In total, 29 thin sections  were produced (thin-section database info), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve, recovered from the 2021 fieldwork; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve, recovered from the 2022 fieldwork; 3) 1 thin-section of a geological sample recovered from previous works (<2021), which was not identified in the latter prospections. Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced, to characterize the chert from the most visited locations. Information about the thin sections, including the laboratories in which they were made, can be consulted in the Supplementary materials.

>>>>

---
title: "Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia."
author:
  - name: Joana Belmiro
    email: jfbelmiro@ualg.pt
    institute: ICArEHB
    correspondence: true
  - name: João Cascalheira
    email: 
    institute: ICArEHB
    correspondence: false
  - name: Xavier Terradas
    email: 
    institute: CSIC
    correspondence: false
institute:
  - ICArEHB: Universidade do Algarve
  - CSIC: Bleh
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

library(readr)
library(dplyr)
library(kableExtra)
library(knitr)

@@

Southern and Western Iberia have often been considered key areas to understanding techno-cultural transitions from the Middle Paleolithic to the end of the Upper Paleolithic. As territories located at the western tip of the European continent and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, these have been regarded as some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022]. For this reason, Southernwestern Iberia has frequently been at the center of some of the most debated topics regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017]. A particularly good example is the region's role in possibly being one of the last territories to be occupied by Neanderthal populations right before their complete disappearance [@finlayson2006; @zilhão2017]. Neanderthals occupied the European continent for more than 300.000 years and are thought to have disappeared while modern humans arrived on the territory [@trinkaus1979; @mellars2004; @finlayson2006; @jennings2011; @rogers2017]. The last territory where Neanderthal populations seemed to exist was Southern Iberia, around c. 37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022] . This region is key to understanding how Neanderthals survived until such a later chronology, the degree and types of interaction those populations may have had with modern humans [@finlayson2006; @zilhão2017], and how and why they eventually went extinct [@dalen2012; @melchionna2018]. Another example is Southwestern Iberia's importance in the discussion of Upper Paleolithic technocomplexes transitions. Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011]. The Heinrich Event 2 (HE 2) at the onset of the Last Glacial Maximum (LGM), for example, is a period marked by important social and technological transformations. This climatic event was characterized by abrupt and drastic climatic changes that impacted human behavior all across westernmost Europe [@gamble2004; @sanchezgoñi2010]. The identification of a Proto-Solutrean phase in central and southern Portugal with a very distinct index fossil (the Vale Comprido point), and its direct association with the HE 2, put these regions amongst some of the most important case studies of how environmental dynamics have affected human adaptations during the last glacial. Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017]. Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources. This continuous use would then create long-term regional adaptative structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches. In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012]. These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints. One of such possible Late Pleistocene eco-cultural niches to which this theoretical framework has been previously applied is the archaeological site of Vale Boi. This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of Iberia [@bicho2013; @cascalheira2013; @cascalheira2017]. Several cross-scale complex interactions have been identified, displaying resilient behaviors throughout the Upper Paleolithic maintained by their eco-cultural niche, but also adaptation behaviors motivated by niche diversity, social networks, and climatic changes [@cascalheira2017]. Some of these resilient elements are, for example, the continuous use of strategies like grease rendering and selective hunting patterns, site function, certain lithic technology patterns, and the functional specialization of lithic raw materials. The maintenancee of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016]. The identification of these patterns has been reliant on the great amount of studies that have originated from the archaeological site of Vale Boi [i.e., @bicho2012; @bicho2017; @pereira2016; @manne2012]. A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on macroscopic preliminary results or the differentiation between possibly local and non-local sorts of raw materials. Despite the scarceness of these studies, chert played a significant role in early and later prehistory at Vale Boi [@cascalheira2010; @marreiros2009; @belmiro2020], and across most of Prehistory in Southern Portugal [@zambujo1998; @bicho2003]. As an essential part of late Pleistocene hunter-gatherer adaptations, lithic raw materials have the potential to provide insights into the adaptive strategies of those populations [@féblot-augustins1993; @kuhn1995; @mellars1996], mostly regarding land-use, technological organization, but also cultural and social interactions. In fact, the selection and procurement of raw material has been suggested as a key stage for the technological organization of hunter gatherer groups [@bamforth1997]. Thus, changes in the frequencies of raw materials within the archaeological records may provide evidence for changes in that organization, or even resilience of specific choices, all of which can reflect culturally transmitted preferences within a group of hunter gatherers [@brown1999]. Several models, both formal and informal, have shown that changes of raw material in the archaeological record may reflect different procurement strategies [@binford1985; @binford1979; @gould1985; @gouldsaggers_1985], either related to mobility strategies [@ambrose1990; @kuhn2004; @kuhn1991; @mccall2007] or changes in the availability of raw materials, possibly related to environmental change [@brantingham2003; @brown1999]. Raw materials are also closely linked to the tools produced with them -- the costs of manufacturing technology can be related to the availability of raw materials in the landscape [@torrence1983; @bousman1993; @parry1987; @mackay2011], or the characteristics of the raw materials which might relate to functionality [@mackay2008; @minichillo2006; @stout2002], through the choice of raw materials for specific technologies [@brown1999]. Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999]. In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like Vale Boi, it is necessary to know the landscape and the raw material sources available in the territory. To distinguish between local and non-local raw materials, first it is necessary to understand the characteristics of the local sources and establish a comparative database which can be used for the analysis of an assemblage. This is true for most of the informal models which try to understand why raw materials change in the archaeological record. Furthermore, formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015]. In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022]. In southern Portugal previous efforts have been made to create such a database. The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from Vale Boi based on macroscopic approaches. The creation of LusoLit, a lithotheque currently hosted at the University of Algarve [@pereirab_presenting_2016], and the collection of samples from several outcrops in the region provided a new leap in the study of chert in the region. A few geological studies have also contributed to understand the availability and characteristics of chert in southern Portugal [i.e., @rocha1976; @ribeiro2005]. Nevertheless, these studies are often unpublished or comprehend answers to geological questions, which hamper the comparative use of the data with archaeological assemblages. Furthermore, a large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists. As such, further analysis of the overall variability, location and availability of chert nodules in the Algarve is necessary to start testing behavioural models of land-use and abiotic resource exploitation. Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity and availability of chert raw materials in the southernmost region of Portugal -- the Algarve. Our main goal is to establish a reference for chert raw materials in an understudied region in what regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere. This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterisation of regional cherts. The data presented here is integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.

@@

The Algarve is the southernmost region of Portugal, framed north by the Alentejo region and east by Spain. To the west and south, it is bordered by the Atlantic ocean. It extends for \~130 km E-W and \~50 km N-S and is characterized by a variety of geomorphic sub-regions and geological units, that make this region a complex territory (\@ref(fig:geomap)). On the north sector of the Algarve, the Serra Algarvia is characterized by a mountainous range with a dense hydrographic network, which separates the Algarve from Alentejo. On the south sector, the Litoral is characterized by a flatter, long strip of land, that extends through all of the coastal strip of the Algarve. The Barrocal is nested between the other sub-regions and has a more moderate relief, characterized by carbonated Jurassic formations and important sub-terranean water circulation [@gago2007].

```{r geomap, echo=FALSE, fig.cap="Geological map of the Algarve region, bordered north by the Alentejo region and east by Spain. The map represents the several geological layers and formations, as well geomorphic sub-regions. The Jurassic layers are colored in green. Metadata was obtained through the SNIAmb (Sistema National de Informação de Ambiente)", out.width='100%'}

knitr::include_graphics("../figures/geology_map.png")

Geologically, the Algarve is composed of two main geological units: the South Portuguese Zone (SPZ) and the Algarve basin. The SPZ is located in the north sector of the Algarve, extending up to Alentejo [@fernandes_new_2012]. Its main lithologies are schist, greywacke and quartzite [@fernandes_new_2012; @pereira2016]. The SPZ is overlain unconformably by the Mesozoic sedimentary rocks of the Algarve basin [@fernandes_new_2012]. The basin corresponds to the Mesozoic-Cenozoic sediments that outcrop south of the Algarve, from the westernmost to the easternmost point of the region, and it is associated with the opening of the central Atlantic Ocean and with the eventual oceanic crust formation in the western part of the Tethys sea, between the Algarve and North Africa [@terrinha_bacia_2013]. Paleozoic sedimentation of the basin started in the Triassic and continued thereon. In the Lower Jurassic (Lower Pliensbachian, also regionally known as Carixian) the basin was divided into two sub-basins -- western and oriental sub-basins [@rocha1976; @terrinha_bacia_2013]. The existence of the two sub-basins and the expansion and retraction of the sea created a variety of sedimentation environments, such as external and internal platforms, continental, hemipelagic and deep marine [@terrinha_bacia_2013], as well as moments of sedimentation hiatus. This variability in deposition environments created a variety of Paleozoic sedimentary facies, with moments of more or less homogeneity throughout this period. For example, during the Lower Pliensbachian, in the Lower Jurassic, the sediments in the western sub-basin can be described as marine of external platform, while the sediments of the eastern sub-basin can be described as marine of internal platform. During the Upper Jurassic however, the basin is marked by a moment of prominent lithofacies variation, followed by a moment of uniformity in both sub-basins [@terrinha_bacia_2013]. Understanding the Algarve basin is key for raw material studies in the Algarve, especially when studying chert, since it is in the basin, more specifically in the Paleozoic sediments, where chert primarily outcrops in the region.

@@

The presence of chert is commonly associated with carbonates in limestone and dolomite formations. This is explained by characteristics (such as the presence of water or specific temperatures and pH) that are ideal for both the formation of limestone and the precipitation of silica [@luedtke1992]. The pelagic and marine environments of the Algarve basin during the Paleozoic gathered those such ideal characteristics, allowing the sedimentation and precipitation of limestone and silica, creating limestone layers with chert nodules. The existence of two basins with different sedimentation environments also shows potential for the precipitation of different types of chert throughout the basin and their differentiation. For example, during sedimentation, skeletal grains of fossils may be preserved. Many of these fossils are restricted to specific sedimentation environments and time intervals [@flügel2010], which may allow the identification of chert outcrops through the fossil content. The basin and sub-basins thus show potential for the existence of different geological formations with different chert types and their study. Previous works, both geological and archaeological, confirm that chert is present in the Algarve in the Jurassic limestone or dolomitic limestone layers of the Algarve basin. This means that chert outcrops can be identified in the central/south sector of the region, from west to east. Variability in chert availability, as well as chert types, is expected, considering that during the sedimentation process, the Algarve basin was already sub-divided and in constant environmental change. Due to this, several formations with chert nodules can be identified in the Algarve, corresponding to different sub-periods of the Jurassic. In the western sector of the Algarve, chert can be found in the Lower Jurassic limestone or dolomitic limestone layers (from the regional Carixian), often visible in areas where the layers are exposed, such as beach-generated cliffs and associated deposits. This includes outcrops such as Cabo de S. Vicente (CSV) and Praia do Belixe (PBLX). These sediments and chert nodules are also available inland, albeit more scarcely, as is the case of the small outcrop named Ferrel, 3 km from the current coastline. Lower Jurassic chert-bearing outcrops are barely existent in the center/east sector of the Algarve, with one single formation with micro-nodules identified in geological works [@oliveira1992]. Middle Jurassic geological layers with chert nodules are only found in the center/east sector of the Algarve in a geological formation called the Malhão Formation. The formation can be described as carbonated, from a marine sedimentation environment . Chert in this formation has been identified in two distinct layers: conglomerates with micritic limestone intercalations with chert beds and nodules, characterized by the presence of sponge spicules and radiolarians [@manuppela1987]; microcrystalline limestones with chert nodules characterized by the presence of silicified malacofauna and silicified corals [@manuppella2007]. Finally, Upper Jurassic sediments with chert nodules have also been mostly identified in the center/east sector of the Algarve, attributed to the Jordana Formation. This formation is characterized by dark-gray limestones, with frequent secondary silicifications with abundant fossil fragments [@manuppela1987; @rocha1989; @manuppella2007]. One single outcrop from the Upper Jurassic has been identified in the western sector of the Algarve basin, located at Praia da Mareta.

@@

To locate and characterize chert outcrops in southern Portugal and understand chert's characteristics, a macroscopic and petrographic approach was applied to the study of geological samples which were collected through fieldwork. A combination of methods in a multilayered approach is an important tool to reconstruct the geological and geographical origin of raw materials, especially since different methods have their inherent limitations. By combining these two different methodologies the goal was to create several layers of analysis, to which new layers may be added if necessary [@brandl2016]. Several other similar methodologies and approaches have been applied in other regions [i.e., @garcía-simón2016; @terradas2017; @tomasso2019; @desoler2020]. However, the chosen analysis techniques should be adapted to the specific geographic context, the research questions, the problematics, and the characteristics of the types of cherts in question [@luedtke1992]. Since only preliminary studies of raw materials were applied in the western portion of southern Portugal, and petrographic data has been shown to provide good results for the characterization of cherts in this region [@ribeiro2005], the two methodologies were chosen for the study. The geological samples used in this study were obtained during fieldwork, between August 2021 and June 2022. The locations prospected were chosen after reviewing previously known research, which included preliminary raw materials studies in the region [@verissimo2004; @pereira2016], geological scientific papers and theses focusing on the Algarve basin and concerning chert outcrops [@rocha1976; @marques1983; @ramalho1985; @ribeiro2005], and geological maps, which signaled the presence of chert nodules within the outcrops[@rocha1979; @rocha1983; @rocha1989; @oliveira1984; @oliveira1992; @manuppela1987; @manuppella2007]. Unpublished data and coordinates for unsurveyed locations with potential for chert-bearing outcrops gathered during the organization of the LusoLit lithotheque were also checked. Whenever coordinates or specific locations for known outcrops were available, these were directly visited and the surrounding area was surveyed to understand the extension of the outcrops and to locate possible secondary deposition outcrops nearby. When no specific locations were described (for example, in geological maps) several areas with more potential to find chert outcrops within one larger formation were surveyed. Samples were collected whenever possible, focusing on both primary and secondary outcrops. When chert nodules within one single outcrop showed macroscopic differences (such as differences in the color, texture, translucency, or cortex), samples of each different nodule were collected, to cover all chert variability within the outcrop. This variability was also recorded using a database (to distinguish between homogeneous or heterogeneous chert nodules within the outcrop) and through photography (\@ref(fig:fieldphoto)). All samples were registered with resource to a free android app (Archaeosurvey) which was designed for archaeological surveys, and records site location and characteristics [@cascalheira2017a], further adapted for raw material source surveys [@abrunhosa2017]. The version of the software used for fieldwork is an adaption of the latter apps and records data related to outcrop characteristics and conditions (i.e. abundance, visibility, access, geomorphology, chert morphology, and conditions). All data related to the app and fieldwork can be found in the Supplementary Online Materials (SOM). Individual IDs were associated with each sample, which includes sequential numbers (based on recovery order, i.e., SP10) and outcrop code (i.e., PdA).

```{r fieldphoto, fig.cap="Sample SP65_MALH from the Malhão outcrop, collected during fieldwork. Left – chert samples collected and nodule still embedded in the parent rock. All chert macroscopic variability was recoded through photography upon sample collection. Right – general photo of the outcrop. The outcrops and points of reference were recorded for contextualization and easiness of future access.", out.width='100%'}

knitr::include_graphics("../figures/fieldwork.png")
